Using Visual Analytics to Explain\\[1ex] Black-Box Machine Learning

### 2018-04-26 tex/abstract.tex
typo fixing
thanks clement

% \begin{abstract}
As machine learning models increase in complexity, the human ability of understanding and interpreting decisions made by those models has not been able to keep up.
The usefulness of white-box analysis techniques, exposing the internal state of models, are limited to relatively simple models and have trouble with complex models, such as deep neural networks.
Recently, black-box machine learning analysis techniques offer model independent insights into the decision making process of machine learning models.
For quickly and effectively gaining insights from those techniques, visual analytics emerges as a powerful set of tools.
We use visual analytics to explore both global and local means of explaining and understanding predictive models via black-box techniques.
We then propose a model diagnostic workflow that uses aggregated instance-level explanations to overcome problems of fully global or local methods.
That is, by avoiding global aggregates, finer details of the decision making process are retained, while going beyond individual instances, analysts are not overwhelmed by the quantity of instances to inspect.
Finally, we show that the Model Diagnostic workflow can not only help improving the models themselves but offers insights about flaws in the \emph{input data}, thus helping with the task of feature engineering.

>>>>

% \begin{abstract}
As machine learning models increase in complexity, the human ability to understand and interpret decisions made by those models has not been able to keep up.
The usefulness of white-box analysis techniques, exposing the internal state of models, is limited to relatively simple models and has trouble with complex models, such as deep neural networks.
Recently, black-box machine learning analysis techniques offer model independent insights into the decision making process of machine learning models.
In order to quickly and effectively gain insights from those techniques, visual analytics emerges as a powerful set of tools.
We use visual analytics to explore both global and local means of explaining and understanding predictive models via black-box techniques.
We then propose the Model Diagnostic workflow that uses aggregated instance-level explanations to overcome problems of fully global or local methods.
That is, by avoiding global aggregates, finer details of the decision making process are retained, while going beyond individual instances, analysts are not overwhelmed by the quantity of instances to inspect.
Finally, we show that the Model Diagnostic workflow can not only help improving the models themselves but offers insights about flaws in the \emph{input data}, thus helping with the task of feature engineering.

### 2018-04-26 tex/introduction/contribution.tex
thanks clement

\subsection{\emph{\prospector}}
In the second stage, we explore the influence of features in the decision making process of predictive models through a technique called partial dependence. Partial dependence computes feature impact on model outcomes by systematically probing the model with artificial inputs. That is, while keeping the rest of the input values the same, the inspected feature assumes all possible values showing the relationship of feature values to the prediction score. Aggregated over all observed instances, the general behavior of the model with respect to a given feature can be inferred. By computing a novel feature importance score from partial dependence relationships unusual and interesting relations can be found quickly without having to inspect all, often several thousand, features.

@@

\subsection{Model Diagnostic Workflow}
In the third stage, we overcome those limitations by leveraging instance-level explanations. Instance-level explanations are defined as the smallest change, to an instance vector, necessary to change the predicted outcome label. Aggregating over those explanations and statistically analyzing the resulting instance subsets, proves as powerful novel approach for understanding the behavior of a model. Additionally, we propose a Model Diagnostic workflow (see \figref{figs:workflow_intro}) that helps identify flaws in the \emph{input data}, used to train and test the model.

>>>>

\subsection{\emph{\prospector}}
In the second stage, we explore the influence of features in the decision making process of predictive models through a technique called \emph{partial dependence}. Partial dependence computes feature impact on model outcomes by systematically probing the model with artificial inputs. That is, while keeping the rest of the input values the same, the inspected feature assumes all possible values showing the relationship of feature values to the prediction score. Aggregated over all observed instances, the general behavior of the model with respect to a given feature can be inferred. By computing a novel feature importance score from partial dependence relationships unusual and interesting relations can be found quickly without having to inspect all, often several thousand, features.

@@

\subsection{Model Diagnostic Workflow}
In the third stage, we overcome those limitations by leveraging \emph{instance-level explanations}. Instance-level explanations are defined as the smallest change, to an instance vector, necessary to change the predicted outcome label. Aggregating over those explanations and statistically analyzing the resulting instance subsets, proves as powerful novel approach for understanding the behavior of a model. Additionally, we propose a Model Diagnostic workflow (see \figref{figs:workflow_intro}) that helps identify flaws in the \emph{input data}, used to train and test the model.

### 2018-04-26 tex/introduction/introduction.tex
thanks clement

With the growing complexity of machine learning models and the need for algorithm agnostic approaches, black-box explanations offer insights into the external behavior of such models.
Visual analytics can be used to leverage the full potential of those explanations in a scalable way.

>>>>

With the growing complexity of machine learning models and the need for algorithm-agnostic approaches, black-box explanations offer insights into the external behavior of such models.
Visual analytics can be used to leverage the full potential of those explanations in a scalable way.

### 2018-04-26 tex/introduction/machine_learning.tex
thanks clement

All of those measures can be read from a confusion matrix, as shown in Figure~\ref{figs:cm_intro}.

@@

\subsubsection{Ensemble Methods}
A powerful method of utilizing several weak models are Ensemble Models.
Instead of taking the prediction of one model, Ensembles combine the prediction of many models in order to get a definite result.
This is typically done either by using the average prediction of multiple models or by majority voting, \ie, the most common prediction among the individual models is used as final prediction \cite{mlbook}.

The main advantage of Ensemble Models is that it can improve the performance of otherwise worse performing models.
Given a set of independent / uncorrelated models the error reduction of the Ensemble is proportional to the number of individual models \cite{dlbook}. % chapter 7.11

@@

Other common layers are ReLU (Rectified Linear Unit) and Softmax which both require $N = M$.
ReLU computes as $y_j(x) = \max{(0, x_j)}$ which basically acts as filter to let only positive values through.
Softmax computes as $y_j(x) = e^{x_j} / \sum^{N}_{i = 1} e^{x_i}$ and scales the input vector in a way that the output vector sums to one.

@@

Training a MLP or any Neural Network works as follows.
We initialize all weights with random numbers.
%\footnote{Other initialization methods such as Xavier also exist.}
For each instance of the training data set, we first compute the outcome with the current weights.
Then we compute the gradient of the loss function, (which can be either the negative log likelihood of training data labels for classification or the mean of squared errors for regression tasks), with respect to each parameter.
The gradient is computed on the predicted outcome $y(x)$ (\ie, values of the final layer) and the \emph{desired} outcome $\hat{y}(x)$ (\ie, value of desired class set to one and zero otherwise).
Then, going in reverse order we numerically compute this gradient for each layer and nudge the weights of those layers into the direction of the gradient.
This procedure is called back-propagation.
Ideally, we repeat back-propagation until all weights converge towards fixed numbers, at which point the model is trained.

Neural networks are a powerful machine learning tool, as they enable any function to be modeled \cite{hornik1991251} and allow for great flexibility through combining different layers.
Multi-Layer Perceptrons can potentially be used on our running Diabetes example from above, as they do not have the limitations of any of the previously discussed models.
However, Neural Networks typically require many thousand training examples in order to properly converge towards a performant model, making them a sub-optimal choice for tasks where acquiring data is time-consuming or expensive.

\subsection{Failure Modes}
Machine learning models not always perform perfectly.
In general, for a given data set, failures to do so typically fall into two main categories: under- and over-fitting \cite{mlbook}.

Under-fitting occurs when a model does not have enough \emph{capacity} to fully learn the input data.
For example, a Logistic Regression model will never be able to learn to classify whether a point lies inside or outside of a circle, using Cartesian coordinates as input features.
This is due to the linear nature of the model.
Under-fitting can be prevented by using a more powerful model or increasing the capacity of the current model.

Over-fitting occurs when a model learns its training data set too precisely.
For example, a Decision Tree might create one leaf node for every instance in the training data memorizing the outcomes of the training data in full.
In this case the model likely won't generalize to new, unseen data.
Thus, in order to detect over-fitting, it is common to keep a separate data set, called hold-out, validation, or test data set, at hand that is not used for training the model.
The trained model is then tested against this data set.
If the model performs well on the training data but not on the test data, then the model is over-fitting.
A way to prevent this is to reduce the capacity of the model.
For example, reducing the allowed maximum depth of a Decision Tree model is a good way to reduce its capacity.

>>>>

All of those measures can be read from a \emph{confusion matrix}, as shown in Figure~\ref{figs:cm_intro}.

@@

\subsubsection{Ensemble Methods}
A powerful method of utilizing several weak models are Ensemble Models.
Instead of taking the prediction of one model, Ensembles combine the prediction of many models in order to get a definite result.
This is typically done either by using the average prediction of multiple models or by majority voting, \ie, the most common prediction among the individual models is used as final prediction~\cite{mlbook}.

The main advantage of Ensemble Models is that it can improve the performance of otherwise worse performing models.
Given a set of independent models with uncorrelated errors the resulting error reduction of the Ensemble is proportional to the number of individual models~\cite{dlbook}. % chapter 7.11

@@

Other common layers are ReLU (Rectified Linear Unit) and Softmax which both require $N = M$.
ReLU computes as $y_j(x) = \max{(0, x_j)}$ which acts as filter to let only positive values through.
Softmax computes as $y_j(x) = e^{x_j} / \sum^{N}_{i = 1} e^{x_i}$ and scales the input vector in a way that the output vector sums to one.

@@

Training a MLP or any Neural Network works as follows.
We initialize all weights with random numbers.
%\footnote{Other initialization methods such as Xavier also exist.}
For each instance of the training data set, we first compute the outcome with the current weights.
Then we compute the gradient of the loss function, (which can be either the negative log likelihood of training data labels for classification or the mean of squared errors for regression tasks), with respect to each parameter.
The gradient is computed on the predicted outcome $y(x)$ (\ie, values of the final layer) and the \emph{desired} outcome $\hat{y}(x)$ (\ie, value of desired class set to one and zero otherwise).
Then, going in reverse order we numerically compute this gradient for each layer and nudge the weights of those layers into the direction of the gradient.
This procedure is called \emph{back-propagation}.
Ideally, we repeat back-propagation until all weights converge towards fixed numbers, at which point the model is trained.

Neural networks are a powerful machine learning tool, as they enable any function to be modeled \cite{hornik1991251} and allow for great flexibility through combining different layers.
Multi-Layer Perceptrons can potentially be used on our running Diabetes example from above, as they do not have the limitations of any of the previously discussed models.
However, Neural Networks typically require many thousand training examples in order to properly converge towards a performant model, making them a sub-optimal choice for tasks where acquiring data is time-consuming or expensive.
Additionally, the large number of (hyper-)parameters also makes them difficult to tune.

\subsection{Failure Modes}
Machine learning models do not always perform perfectly.
For a given data set, failures to do so typically fall into two main categories: under- and over-fitting~\cite{mlbook}.

\emph{Under-fitting} occurs when a model does not have enough \emph{capacity} to fully learn the input data.
For example, a Logistic Regression model will never be able to learn to classify whether a point lies inside or outside of a circle, using Cartesian coordinates as input features.
This is due to the linear nature of the model.
Under-fitting can be prevented by using a more powerful model or increasing the capacity of the current model.

\emph{Over-fitting} occurs when a model learns its training data set too precisely.
For example, a Decision Tree might create one leaf node for every instance in the training data memorizing the outcomes of the training data in full.
In this case the model likely will not generalize to new, unseen data.
Thus, in order to detect over-fitting, it is common to keep a separate data set, called \emph{hold-out}, \emph{validation}, or \emph{test} data set, at hand that is not used for training the model.
The trained model is then tested against this data set.
If the model performs well on the training data but not on the test data, then the model is over-fitting.
A way to prevent this is to reduce the capacity of the model.
For example, reducing the allowed maximum depth of a Decision Tree model is a good way to reduce its capacity.

### 2018-04-26 tex/introduction/motivation.tex
thanks clement

The above examples illustrate the need for human understanding and supervision in machine learning.
However, those cases required intricate understanding of the used machine learning models and are not easily generalized to other techniques.
For example, in the case of Caruana~\etal\cite{Caruana:2015:IMH:2783258.2788613}, the incorrect correct result was easily identified since the authors used a special, intelligible, model.
However, not all models used in machine learning are, or even attempt to be, easily interpretable by human experts.
To counteract this, visual analytics has been successfully used to visualize machine learning processes in order to understand and gain insights into machine learning models~\cite{DBLP:journals/corr/StrobeltGHPR16,DBLP:journals/corr/abs-1710-06501,2017arXiv170401942K,DBLP:journals/corr/abs-1710-10777,8019879,8019872,DBLP:journals/tvcg/LiuXLWWZ18,8019878}. 
% Visual analytics is the area of studying interactive graphical representations of abstract information with the help of computational and statistical procedures, in order to effectively analyze complex data.
With regards to understanding the decision making of predictive machine learning models, there are two main strategies: white-box and black-box analysis \cite{class_signatures}.

>>>>

The above examples illustrate the need for human understanding and supervision in machine learning.
However, those cases required intricate understanding of the used machine learning models and are not easily generalized to other techniques.
For example, in the case of Caruana~\etal\cite{Caruana:2015:IMH:2783258.2788613}, the incorrect correct result was easily identified since the authors used a special, intelligible, model.
However, not all models used in machine learning are, or even attempt to be, easily interpretable by human experts.
To counteract this, visual analytics has been successfully used to visualize machine learning processes in order to understand and gain insights into machine learning models~\cite{DBLP:journals/corr/StrobeltGHPR16,DBLP:journals/corr/abs-1710-06501,2017arXiv170401942K,DBLP:journals/corr/abs-1710-10777,8019879,8019872,DBLP:journals/tvcg/LiuXLWWZ18,8019878}. 
% Visual analytics is the area of studying interactive graphical representations of abstract information with the help of computational and statistical procedures, in order to effectively analyze complex data.
With regards to understanding the decision making of predictive machine learning models, there are two main strategies: \emph{white-box} and \emph{black-box} analysis \cite{class_signatures}.
